Towards a Wide-Coverage Grammar for Swedish\\ Using GF

### 2012-01-12 doc/report/report.tex
report corrections

\begin{document}
% ska gå att förstå och locka till läsning
%ska tydligt beskriva vad rapporten handlar om
%ska innehålla vettiga sökord
%TODO page numbers!!
\title{Towards a Wide-Coverage Grammar for Swedish Using GF} %A nice titleScaling up a GF grammar and preparing it for cool stuff for Swedish}
\author{Malin Ahlberg}
\maketitle
\newpage

\abstract{
The thesis describes the work formalising an important fragment of Swedish
in the GF formalism and to compare and evaluate the possibilities of using GF
for modelling large-scale grammars. Grammatical constructions specific to and
frequent in Swedish have been investigated and formalised in the functional
programming language.
%~~achieve a covering Swedish~~ grammar implemented
%TODO first sentence should be more, no understatements!
%in GF -- to model natural language by functional programming. 
The idea has been to combine existing language technology resources with
new techniques, and the aim to obtain a system for analysing written text.
To reach this goal, problems of computational as well as linguistic nature had to be solved.
\\We present the methods and technologies used and further inspects the possibilities
of reaching our long-term goal; to accomplish a parser for unrestricted Swedish, by
combining our rule-based grammar with statistical methods.
\\Our contribution is a wide-coverage GF lexicon, a translation of a Swedish treebank
to GF notation and a Swedish grammar implementation. The grammar is based
on the multilingual abstract syntax given in GF resource library, and now
extends it to cover constructions specific to Swedish.
We further give an example of the advantage of using dependent types when
describing grammar and syntax, in this case for dealing with reflexive pronouns.
%the linguistic phenomena

@@

\newpage
\pagestyle{empty}
\section*{Acknowledgments}
\vspace{5mm}
Many people have helped me during this work and made this thesis possible.
I would first of all like to thank
Center of Language Technology, that has funded the project.

@@

I am also grateful to Krasimir Angelov, Markus Forsberg, Peter Ljunglöf, Lars Borin
and many others who have contributed with ideas and inspiration, and for their
interest in this work.

Finally, I would also like to thank my friends and family. Special thanks to Dan % Rosén
for all his support, advice and patience and  -- most importantly -- for being
such a good friend. %for having listened to 
%my monologues about the daily progress. 
% Ramona, Aarne, Elisabet, Markus, Peter?, Lars, Dan, Krasimir etc
% Funding!

% check references! lncs is good maybe. lnai
\newpage
%\pagestyle{empty}
%\cleardoublepage
\tableofcontents
%
\addtocontents{toc}{\protect\thispagestyle{empty}}
%\pagestyle{empty}
%\cleardoublepage
\newpage
%\pagenumberingstyle{plain}
\setcounter{page}{1} %reset the page counter
\chapter{Introduction}
\thispagestyle{plain}
% more intro to gf. dependently typed...high-level formalism
Grammatical Framework \cite{gfbok} is a dependently typed grammar formalism,
with facilities for multilinguality. It is based on Martin-Löf type theory which allows
reasoning within the programming language.
GF has so far been successfully used for 
controlled languages \cite{cnl}, but recent experiments have showed
that it is also possible to use the framework for parsing free language \cite{patent}.
% >> although the success rate is going steeply down when using uncontrolled nl. Better with annotated.
% --> sentence about connection GF - grammar based parser.
This project focuses on implementing a wide-coverage Swedish grammar, which
describes a parser accepting all sentences that can be expressed by the grammar.

Parsing, that is the task of automatically identifying morphological and
syntactical structures, is subject to increasing interest, especially
considering the steadily growing amount of data available online. 
So far there is no freely available grammar-driven parser that gives a deep
analysis for Swedish. The fact that %the existing rule-based Swedish
a parser or grammar is not freely available does not only restrict its
usage but also its possibilities of being further developed and our
goal is
to create an open source Swedish grammar. % from which we derive a parser
%accepting all sentences described by the given rules.
As a freely available resource, it can be continuously enhanced and
made use of in other projects.\\
To build the grammar from scratch would be not only time consuming but
would also mean that already existing resources would have to be reimplemented.
In order to not reinvent the wheel we proceed from a combination of well-tested
resources.
We start from a GF resource grammar, consisting of an abstract and a concrete syntax
file defining a set of rules describing morphology and syntax.
%TODO verify this, check all occurences of grammar
From now on, this is what we mean by grammar, as
opposed to grammar in the traditional linguistic sense. 
From GF, we get %and thereby get
%The usage of GF allows us to start from
a well-defined system for describing language, as well as a strong connection to
more than 20 other languages implemented in the framework. Further, we use
the extensive lexicon Saldo and the treebank Talbanken.
%>>The grammar defines a parser, which accepts all constructions defined by the grammar.

@@

\section{Aims}
%>> enhance gf...
\protect\thispagestyle{plain}
The purpose of this project has been to prepare the GF grammar for parsing of 
unrestricted Swedish.
This has meant to develop earlier techniques to fit for Swedish, create methods
for achieving and keeping a large-scale lexicon and to adapt the existing
resource grammar to fit more language specific constrictions.
Three sub-goals has been identified:
\begin{itemize}
\item Extending the Swedish GF grammar
\item Importing the lexicon Saldo
\item Creating translation between Talbanken and GF
\end{itemize}
%We hence needed to extend the Swedish grammar to cover
%the language specific constructions. Further a large lexicon was
%needed, as well as a method for evaluation.
%During this project, we have worked on and explored those things and
%while identifying problems and future work.

\section{Outline}
The thesis is divided into 6 chapters. We start by giving some background
information to areas relevant to the project. %Chapter \ref{sec:background} 
Grammatical Framework is presented in section \ref{sec:gf} while a more profound
description of the Swedish resource grammar is given in section \ref{sec:swegf}.
Section \ref{sec:swedish} gives an introduction to the Swedish language, and
brief presentations of Saldo and Talbanken are found in section \ref{sec:saldo} and
\ref{sec:talbanken} respectively.
A summary of related work can be found in section \ref{sec:related}.

@@

GF is a strongly typed functional programming language, inspired by
%programming languages like 
ML \cite{ml} and Haskell \cite{haskell}. It is also a logical framework,
and the built-in functionality for logic and reasoning 
is inspired by \textlambda Prolog \cite{prolog} and 
by Agda \cite{agda}, with which GF also shares its type checking algorithm.
The first version of the framework was implemented at Xerox Research Center
in Grenoble and is now mainly developed in Gothenburg. One of the biggest
achievements is a library covering the 
basic morphological and syntactic structures of more than
20 languages (see section \ref{sec:resources}).

%> dependent and implicit types : agda. (read gf book and krasses book)
A grammar written in GF can be used for both parsing and generation.
The parsing algorithm is incremental and has polynomial time and space
complexity \cite{gfMech}. 
The GF package also provides various tools for working with and using the grammars:
a compiler, an interpreter and a runtime system.
The grammars can be compiled into portable grammar format (PGF) \cite{pgf},
supported by %There are libraries for using PGF files with
Java, Java script, C and Haskell libraries. 
%>> Haskell only complete, also or C\# and C.  %may also be used in Android applications.
The interpreter, the GF shell, allows
the user to test grammars by commands for parsing, visualization of parse trees,
random generation, word alignment, morphological quizzes etc.
An interactive GF editor and the GF shell can be tried out online 
\footnote{http://www.grammaticalframework.org/demos/gfse/}. See figure \ref{fig:shell},
 The first command will parse the given
sentence and visualize the resulting parse tree. The second will visualize the 
abstract syntax tree. The trees will be showed in the pdf-format by the program evince.
\thispagestyle{plain}
\begin{figure}[h]
\begin{verbatim} 
> parse "jag ser katten" | visualize_parse -format=pdf -view=evince
> parse "jag ser katten" | visualize_tree  -format=pdf -view=evince
\end{verbatim}
\caption{Example of how the GF shell can be used.}\label{fig:shellvp}
\label{fig:shell}
\end{figure}

% nice: test vt -format=pdf -view=evince -nofun
The visualization of the trees is done with graphviz\footnote{http://www.graphviz.org/}.
Figure \ref{fig:gftree1} shows the results from running the commands in figure \ref{fig:shellvp}
%A GF parse tree has two visualized representations, the \textit{parse tree} and
%the \textit{abstract tree}.
The \textit{abstract tree} shows which functions GF uses
to parse a sentence.
The \textit{parse tree} (\ref{pic:gfStree}) shows which types
that are assigned to the words and phrases. 
\begin{figure}[h]
\centering
\subfloat[GF abstract tree]{\label{pic:gfAtree}\includegraphics[width=100mm]{gfTree.pdf}}
\subfloat[Swedish parse tree]{\label{pic:gfStree}\includegraphics[width=30mm]{gfSTree.png}}
\caption{Abstract tree and parse tree for the sentence ``Jag ser katten".}
\label{fig:gftree1}
\end{figure}
\vspace{10mm}
% do not know about abstract/concrete yet: and is consequently the same for all concrete syntaxes.
The reader should keep in mind that the visualized parse trees is not a
complete representation. It does not reproduce all information from the abstract tree.
%If the end node of a  branch has no corresponding word, it is excluded from
%the visualized parse tree. This is the case for the 
For example, the definiteness and number of the noun \emph{`katten'}, which is shown as
\verb-DetQuant DefArt NumSg- in the abstract tree.
%and the information \verb-DetQuant DefArt NumSg- is not shown in figure \ref{pic:gfStree}.
In the corresponding English parse tree, figure \ref{pic:gfEtree},
the noun is explicitly quantified by the article \emph{``the"},
and the determiner is therefore shown in the parse tree.\\
%>> how parse trees work, not correct complete representation.
%>> not a constituent structure. compare to lfg, fstructure, cstructure.

@@

%TODO read again
The scope of GF grammars has so far been controlled language, a limited
language used in a restricted domain. By
restricting the coverage, the number of ambiguities
can be limited and we can ensure that the semantics is preserved during
translation. Inherited ambiguities as well as ambiguities arising from multilinguality will
remain, but can be controlled more easily. %Example of domainspecific word?

The use of controlled language thus gives a possibility of good and reliable translation
and GF has successfully been used in project like %Some examples of projects using GF are 
WebAlt \cite{webalt}, a project which aims to develop language independent
material for mathematics problems. The formal software verification tool
KeY \cite{key}, in which GF is used for translating specification to natural language.
%>> key, talk, hats, bom (see references in Ramonas thesis).
%>> use talk here and move key to other place??
GF can be used for describe dialog grammars see \cite{talk}, and
the framework is further used in the
European project MOLTO \footnote{http://www.molto-project.eu/} for online
translation between up to 15 language.\\

@@

%>> move much of this to Swedish GF part?
Figure \ref{fig:gfSweCnc1} shows how the abstract grammar can be implemented
for Swedish. Nouns, verbs and sentences are all defined as strings, \verb|Str|.
The function \verb|Pred| simply glues the two strings ``katten" and ``sover" together:\\
\verb|Pred cat sleep = "katten sover"|.\\
We get a more complicated example if we allow the nouns to be used in both
plural and singular. We add a category \verb|N'| to the
abstract, which represents a noun with a fixed number,
%Note that \verb-N'- is normally used for nouns with a fixed definitness, but for 
%now we concentrate on the number only. Hence \verb|N| now means a noun in any of the number,
and we introduce two functions that set the number: \verb|NSg : N -> N'| and
\verb|NPl : N -> N'|.
\begin{figure}[h!]
\begin{verbatim}              
 abstract TestGrammar = {          concrete TestGrammarSwe of TestGrammar = {
  cat N ; N' ; V ; S ;               lincat V, S, N' = Str ;
                                            N = {s : Num => Str} ;
  fun                                lin   
    Pred : N' -> V -> S ;              Pred n v = n ++ v ;
    NSg : N -> N' ;                    NPl n = n.s ! Pl ;
    NPl : N -> N' ;                    NSg n = n.s ! Sg ;
    cat_N : N ;                        cat_N = {s = table {Sg => "katten" ;
    sleep_V : V ;                                          Pl => "katterna"}};
 }                                     sleep_V = "sover"  ;
                                     param Num = Sg | Pl ;
                                     }
\end{verbatim}           
\caption{A modified grammar}
\label{fig:gfTest2}
\end{figure}
Figure \ref{fig:gfTest2} introduces some new concepts: records, tables and parameters.
In the concrete syntax, \verb|N| is defined to be a record consisting of the field
\verb|s|. The type of \verb|s|, \verb-Num => Str- shows that it is a table, which given a
parameter of type \verb|Num| returns a string. \verb|Num| is defined 
%as a parameter, which can
to either have value \verb|Sg| or \verb|Pl|. 
The dot (\verb-.-) is used for projection and the bang (\verb-!-) as a selection operator.
%In \verb|NPl| and \verb|NSg|,
\verb|n.s ! Sg| thus means that we use the branch for \verb-Sg- in
field \verb|s| of \verb|n|.
%, and then the selection operator \verb|!| is used to select a branch in the table.

@@

\subsection{The resource library}
\label{sec:resources}
%>>Best part? Biggest library?
%>>Most general resource, first large grammar, covers fundamental syntactics.
The GF package provides an useful resource library \cite{gf-resource}, covering the
fundamental morphology and syntax of more than 20 languages.
There is also a small test lexicon included, containing a few hundred common
words.
%is a very useful part of the framework. It is provided in the 
%, they give a good start for all who want to write an application grammar.
%of a language for each application, all
%this can be found as libraries. So far there are 20 languages in the library, and
The languages all share the same abstract syntax which is  valuable
%they constitute  a very useful resource 
when implementing multilingual application grammars.

The resource grammars describe how to construct phrases and sentences and how to
decline words. The latter is done by smart paradigms: functions that analyse
some given forms of a word to find out how the inflection table should look.
For example, the declination of many Swedish nouns can be determined by looking
at the singular form only. 
\begin{wrapfigure}{l}{0.4\textwidth}
\begin{tabular}{| l | l |}
\hline
1st declination & 5th declination \\
\hline
flick\textbf{a}    &     hjärt\textbf{a}   \\
flick\textbf{an}    &    hjärt\textbf{at}  \\
flick\textbf{or}    &    hjärt\textbf{an}  \\
flick\textbf{orna}  &    hjärt\textbf{ana} \\
\hline
\end{tabular}
\end{wrapfigure}
This is the case for ``flicka", a noun belonging to the first declination. %the correct declination
%can be guessed by using the first declination. % and noting that the word ends with an ``a". 
%Like most Swedish nouns ending with an ``a", it follows the first declination.
For others, like "hjärta", also the plural form "hjärtan" is needed.
%, even though it ends with an ``a". 
The worst case is nouns that need four forms, both singular and
plural in definite and indefinite form.
Section \ref{sec:swegf} will give a more thorough description of the Swedish resource grammar.

@@

\section{Talbanken}
\label{sec:talbanken}
For testing and evaluation of the grammar and lexicon, we needed to be able to
compare them against a reliable source.
Talbanken \cite{talbanken} was perfect for our purpose,
being a freely available, manually annotated, large-scale treebank.
%It was created in the 1970s at Lund University.
It is analyzed with the MAMBA annotation scheme (Teleman, 1974) and 
consists of four
parts. Two of them are transcriptions of spoken language, one a collection of
text written by high school students, and one, section P,
consists of professionally written Swedish gathered from newspapers, brochures and textbooks.
\\
Talbanken was also used to train the Swedish version of the Malt parser \cite{malt}
and was then redistributed in an updated version,
Talbanken05 \cite{talbanken05}.
%This work was done by Nivre, Hall and Nilsson in 2005.
It is released in Malt\footnote{http://w3.msi.vxu.se/~nivre/research/MaltXML.html} 
and Tiger\footnote{http://www.ims.uni-stuttgart.de/projekte/TIGER/TIGERSearch/doc/html/TigerXML.html}
XML-formats
where the trees had been made deeper and more detailed while still containing
the lexical MAMBA layer. \\
The Malt parser was trained on section P of Talbanken, and these
more than 6000 sentences have been used our project. 
The treebank has served as an inspiration and an evaluation source throughout the
project. An automatic mapping between its trees and the abstract trees from GF has been
done, which will be explained in section \ref{sec:Mapping}.
%Differences in analyse, do we want a similar? 
%Uses of Talbanken in project, mapping to evaluate and test.  
%Look at output from Maltparser, could also comparing our parse trees to this.

\section{Saldo}
\label{sec:saldo}
A good parser needs a good lexicon. We have used Saldo \cite{saldo}, a
large electronic lexicon developed and maintained at Gothenburg University. It is
built on Svenskt Associationslexikon and contains information about more than 
120 000 modern Swedish words.
For each word there is semantic, syntactical and a morphological
information. The user can find examples of usage in corpora,
graphs of semantically connected words and some suggestions for how to analyse compounds. \\
\begin{figure}[h!]
\centering
\subfloat[Morphological info for ``katt"]{\label{pic:saldoTab}\includegraphics[width=40mm]{saldotab.png}}
\hspace{20mm}
\subfloat[Graph showing the hyponyms of ``katt"]{\label{pic:saldoWheel}\includegraphics[width=70mm]{saldograph.png}}
\caption{Information in Saldo}
\label{fig:saldo}
\end{figure}

>>>>

\begin{document}
% ska gå att förstå och locka till läsning
%ska tydligt beskriva vad rapporten handlar om
%ska innehålla vettiga sökord
\title{Towards a Wide-Coverage Grammar for Swedish\\ Using GF} %A nice titleScaling up a GF grammar and preparing it for cool stuff for Swedish}
\author{Malin Ahlberg}
\maketitle
\newpage

\abstract{
The goal of this thesis is to enable open-domain parsing for Swedish
by using the GF \\grammar formalism,
%by %The idea has been to% 
combining existing language technology resources with
new techniques.
The work includes the development of a covering grammar,
%in the GF formalism
to identify and formalize \\
grammatical constructions %specific to and
frequent in Swedish. % have been investigated and formalised in the functional
%programming language.
We compare and evaluate the possibilities of using GF
for modelling large-scale grammars and for 
%and the aim to reaching our goal;
obtaining a system for analysing written text.
To reach this goal, problems of computational as well as linguistic nature had to be solved.
%~~achieve a covering Swedish~~ grammar implemented
%TODO first sentence should be more, no understatements!
%in GF -- to model natural language by functional programming. 
\\We present the methods and technologies used and further inspect the possibilities
of reaching our long-term goal; to accomplish a parser for unrestricted Swedish, by
combining our rule-based grammar with statistical methods.
\\Our contribution is a wide-coverage GF lexicon, a translation of a Swedish treebank
to GF notation and an extended Swedish grammar implementation. The grammar is based
on the multilingual abstract syntax given in GF resource library, and now
also covers constructions specific to Swedish.
We further give an example of the advantage of using dependent types when
describing grammar and syntax, in this case for dealing with reflexive pronouns.
%the linguistic phenomena

@@

\newpage
\pagestyle{empty}
\section*{Acknowledgments}
\vspace{5mm}
Many people have helped me during this project and made work thesis possible.
I would first of all like to thank
Center of Language Technology, that has funded the project.

@@

I am also grateful to Krasimir Angelov, Markus Forsberg, Peter Ljunglöf, Lars Borin
and many others who have contributed with ideas and inspiration and shown %for their
interest in this work.

Finally, I would like to thank my friends and family. Special thanks to Dan % Rosén
for all his support, advice and patience and  -- most importantly -- for being
such a good friend. %for having listened to 
%my monologues about the daily progress. 
% Ramona, Aarne, Elisabet, Markus, Peter?, Lars, Dan, Krasimir etc
% Funding!

% check references! lncs is good maybe. lnai
\newpage
%\pagestyle{empty}
%\cleardoublepage
\tableofcontents
%
\addtocontents{toc}{\protect\thispagestyle{empty}}
%\pagestyle{empty}
%\cleardoublepage
\newpage
%\pagenumberingstyle{plain}
\setcounter{page}{1} %reset the page counter
\chapter{Introduction}
\thispagestyle{plain}
% more intro to gf. dependently typed...high-level formalism
Grammatical Framework \cite{gfbok} is a dependently typed grammar formalism.
%with facilities for multilinguality. 
It is based on Martin-Löf type theory which allows
reasoning within the programming language.
GF has strong support for multilinguality and has so far been successfully used for 
controlled languages \cite{cnl}. Recent experiments have showed
that it is also possible to use the framework for parsing free language \cite{patent}.
% >> although the success rate is going steeply down when using uncontrolled nl. Better with annotated.
% --> sentence about connection GF - grammar based parser.
%This project focuses on implementing a wide-coverage Swedish grammar, which
%describes a parser accepting all sentences that can be expressed by the grammar.
Parsing, that is the task of automatically identifying morphological and
syntactical structures, is subject to increasing interest, especially
considering the steadily growing amount of data available online. 
So far there is no freely available grammar-driven parser that gives a deep
analysis for Swedish. The fact that %the existing rule-based Swedish
a parser or grammar is not freely available does not only restrict its
usage but also its possibilities of being further developed. Our
goal is
to create an open-source Swedish grammar from which we derive a parser
accepting all sentences described by the given rules.
As a freely available resource, it can be continuously enhanced and
made use of in other projects.\\

To build the grammar from scratch would be not only time consuming but
would also mean that already existing systems would have to be reimplemented.
In order to not reinvent the wheel we proceed from a combination of well-tested
sources.
We start from a GF resource grammar, consisting of an abstract and a concrete syntax
file defining a set of rules for morphology and syntax.
%TODO verify this, check all occurences of grammar
This is what is meant by grammar in this thesis, as
%From now on, this is what we mean by grammar, as
opposed to grammar in the traditional linguistic sense. 
From GF, we get %and thereby get
%The usage of GF allows us to start from
a well-defined system for describing language, as well as a strong connection
to and possibility of translation between the 
more than 20 other languages implemented in the framework. Further, we use
the extensive lexicon Saldo and the treebank Talbanken.
%>>The grammar defines a parser, which accepts all constructions defined by the grammar.

@@

\section{Aims}
%>> enhance gf...
\protect\thispagestyle{plain}
The purpose of this project has been to prepare the GF grammar for parsing of 
unrestricted Swedish.
This has meant to develop earlier techniques to fit for Swedish, create methods
for achieving and keeping a large-scale lexicon and to adapt the existing
resource grammar to model more language specific constrictions.
The project was divided into three subsections, aiming at
%Three sub-goals has been identified:
\begin{itemize}
\item Extending the Swedish GF grammar
\item Importing the lexicon Saldo
\item Creating translation between Talbanken and GF
\end{itemize}
%We hence needed to extend the Swedish grammar to cover
%the language specific constructions. Further a large lexicon was
%needed, as well as a method for evaluation.
%During this project, we have worked on and explored those things and
%while identifying problems and future work.

\section{Outline}
The thesis is divided into 6 chapters. We start by giving background
information of areas relevant to the project. %Chapter \ref{sec:background} 
Grammatical Framework is presented in section \ref{sec:gf} while a more profound
description of the Swedish resource grammar is given in section \ref{sec:swegf}.
Section \ref{sec:swedish} gives an introduction to the Swedish language, and
brief presentations of Saldo and Talbanken are found in section \ref{sec:saldo} and
\ref{sec:talbanken} respectively.
A summary of related work can be found in section \ref{sec:related}.

@@

GF is a strongly typed functional programming language, inspired by
%programming languages like 
ML \\\cite{ml} and Haskell \cite{haskell}. It is also a logical framework,
and the built-in functionality for logic and reasoning 
is inspired by \textlambda Prolog \cite{prolog} and 
by Agda \cite{agda}, with which GF also shares its type checking algorithm.
The first version of the framework was implemented at Xerox Research Center
in Grenoble and is now mainly developed in Gothenburg. One of the biggest
achievements is a library covering the 
basic morphological and syntactic structures of more than
20 languages (see section \ref{sec:resources}).

%> dependent and implicit types : agda. (read gf book and krasses book)
A grammar written in GF can be used for both parsing and generation.
The parsing algorithm is incremental and has polynomial time and space
complexity \cite{gfMech}. 
The GF package also provides various tools for working with and using the grammars:
a compiler, an interpreter and a runtime system.
The grammars can be compiled into portable grammar format (PGF) \cite{pgf},
supported by %There are libraries for using PGF files with
Java, Java script, C and Haskell libraries. 
%>> Haskell only complete, also or C\# and C.  %may also be used in Android applications.
The interpreter, the GF shell, allows
the user to test grammars by commands for parsing, visualization of parse trees,
random generation, word alignment, morphological quizzes etc.
The shell can be tried out online 
together with an interactive GF editor %and the GF shell 
\footnote{http://www.grammaticalframework.org/demos/gfse/}.
\thispagestyle{plain}
\begin{figure}[h]
\begin{verbatim} 
> parse "jag ser katten" | visualize_tree  -format=pdf -view=evince
> parse "jag ser katten" | visualize_parse -format=pdf -view=evince
\end{verbatim}
\caption{Example of how the GF shell can be used.}\label{fig:shellvp}
\label{fig:shell}
\end{figure}
Figure \ref{fig:shell} shows commands for parsing, where the results are shown as
in figure \ref{fig:gftree1}. 
GF uses Graphviz\footnote{http://www.graphviz.org/}
to visualize the trees and the given commands (fig. \ref{fig:shell}) specify that the
output format should be PDF-files and that these should be opened by the program
Evince. The user can choose to see the \textit{parse tree} (fig. \ref{pic:gfStree})
or the \textit{abstract tree} (fig. \ref{pic:gfAtree}). Abstract trees are more verbose and show all
functions and types used for parsing the sentence. 
\begin{figure}[h]
\centering
\subfloat[Swedish parse tree]{\label{pic:gfStree}\includegraphics[width=30mm]{gfSTree.png}}
\subfloat[GF abstract tree]{\label{pic:gfAtree}\includegraphics[width=100mm]{gfTree.pdf}}
\caption{Abstract tree and parse tree for the sentence ``Jag ser katten".}
\label{fig:gftree1}
\end{figure}
\vspace{10mm}

Parse trees on the other hand show only the types assigned to the words and phrases.
Information about tense, polarity etc. which are explicitly given in the abstract tree are
not reproduced in the parse tree. Hence, parse trees do not give complete representations
but model the parse resluts in a transparent manner. For our example, the definiteness and
number of the noun `\emph{katten}' is shown as \verb-DetQuant DefArt NumSg- in
the abstract tree while the parse tree only show that a the noun phrase consists of one noun. 
% is used but not in the parse tree.
In the corresponding English parse tree, figure \ref{pic:gfEtree},
the noun is explicitly quantified by the article \emph{`the'},
and the determiner, the first argument to the function \verb-DetCN-, is
therefore shown in the parse tree.\\

% The first command will parse the given
%sentence and visualize the resulting parse tree. The second will visualize the 
%abstract syntax tree. The trees will be showed in the pdf-format by the program evince.
%% nice: test vt -format=pdf -view=evince -nofun
%The visualization of the trees is done with graphviz\footnote{http://www.graphviz.org/}.
%Figure \ref{fig:gftree1} shows the results from running the commands in figure \ref{fig:shellvp}
%%A GF parse tree has two visualized representations, the \textit{parse tree} and
%%the \textit{abstract tree}.
%The \textit{abstract tree} shows which functions GF uses
%to parse a sentence.
%The \textit{parse tree} (\ref{pic:gfStree}) shows which types
%that are assigned to the words and phrases. 
%% do not know about abstract/concrete yet: and is consequently the same for all concrete syntaxes.
%The reader should keep in mind that the visualized parse trees is not a
%complete representation. It does not reproduce all information from the abstract tree.
%%If the end node of a  branch has no corresponding word, it is excluded from
%%the visualized parse tree. This is the case for the 
%For example, the definiteness and number of the noun \emph{`katten'}, which is shown as
%\verb-DetQuant DefArt NumSg- in the abstract tree.
%%and the information \verb-DetQuant DefArt NumSg- is not shown in figure \ref{pic:gfStree}.
%In the corresponding English parse tree, figure \ref{pic:gfEtree},
%the noun is explicitly quantified by the article \emph{`the'},
%and the determiner is therefore shown in the parse tree.\\
%>> how parse trees work, not correct complete representation.
%>> not a constituent structure. compare to lfg, fstructure, cstructure.

@@

%TODO read again
The scope of GF grammars has so far been controlled language, a subset of a
natural language used for a restricted domain. By
restricting the coverage, the number of ambiguities
can be limited and it can be ensured that the semantics is preserved during
translation. Inherited ambiguities as well as ambiguities arising from multilinguality will
remain, but can be controlled more easily. %Example of domainspecific word?

The use of controlled language thus gives the possibility of good and reliable
translation and GF has successfully been used for this in a number of projects.
%like %Some examples of projects using GF are 
WebAlt \cite{webalt} aims to develop language independent
material for mathematics problems by using GF grammars.
The formal software verification tool
KeY \cite{key} used GF for translating formal specification to
natural language.
%>> key, talk, hats, bom (see references in Ramonas thesis).
%>> use talk here and move key to other place??
GF has further been used for describing dialogue grammars \cite{talk}, and
the framework is one of the main component in the
European project MOLTO \footnote{http://www.molto-project.eu/}, which develop
online translation between to 15 languages.\\

@@

%>> move much of this to Swedish GF part?
Figure \ref{fig:gfSweCnc1} shows how the abstract grammar can be implemented
for Swedish. Nouns, verbs and sentences are all defined as strings, \verb|Str|.
The function \verb|Pred| simply glues the two strings `katten' and `sover' together:\\
\verb|Pred cat sleep = "katten sover"|.\\
We get a more complicated example if we allow the nouns to be used in both
plural and singular. We add a category \verb|N'| to the
abstract, which represents a noun with a fixed number,
%Note that \verb-N'- is normally used for nouns with a fixed definitness, but for 
%now we concentrate on the number only. Hence \verb|N| now means a noun in any of the number,
and we introduce two functions for setting the number: \verb|NSg : N -> N'| and
\verb|NPl : N -> N'|.
\begin{figure}[h!]
\begin{verbatim}              
 abstract TestGrammar = {          concrete TestGrammarSwe of TestGrammar = {
  cat N ; N' ; V ; S ;               lincat V, S, N' = Str ;
                                            N = {s : Num => Str} ;
  fun                                lin   
    Pred : N' -> V -> S ;              Pred n v = n ++ v ;
    NSg : N -> N' ;                    NPl n = n.s ! Pl ;
    NPl : N -> N' ;                    NSg n = n.s ! Sg ;
    cat_N : N ;                        cat_N = {s = table {Sg => "katten" ;
    sleep_V : V ;                                          Pl => "katterna"}};
 }                                     sleep_V = "sover"  ;
                                     param Num = Sg | Pl ;
                                     }
\end{verbatim}           
\caption{A modified grammar}
\label{fig:gfTest2}
\end{figure}
Figure \ref{fig:gfTest2} introduces some new concepts: records, tables and parameters.
In the concrete syntax, \verb|N| is defined to be a record consisting of the field
\verb|s|. The type of \verb|s|, \verb-Num => Str- shows that it is a table, which given a
parameter of type \verb|Num| returns a string. \verb|Num| is defined 
%as a parameter, which can
to either have value \verb|Sg| or \verb|Pl|. 
The dot (\verb-.-) is used for projection and the bang (\verb-!-) as a selection operator.
%In \verb|NPl| and \verb|NSg|,
\verb|n.s ! Sg| thus means that we use the branch for \verb-Sg- in
field \verb|s| of \verb|n|.
%, and then the selection operator \verb|!| is used to select a branch in the table.

@@

\subsection{The resource library}
\label{sec:resources}
%>>Best part? Biggest library?
%>>Most general resource, first large grammar, covers fundamental syntactics.
The GF package provides an useful resource library \cite{gf-resource}, covering the
fundamental morphology and syntax of more than 20 languages.
There is also a small test lexicon included, containing a few hundred common
words.
%is a very useful part of the framework. It is provided in the 
%, they give a good start for all who want to write an application grammar.
%of a language for each application, all
%this can be found as libraries. So far there are 20 languages in the library, and
Since the languages all share the same abstract syntax translation between any given
pair is possible, which is valuable
%they constitute  a very useful resource 
when implementing multilingual applications.

The resource grammars describe how to construct phrases and sentences and how to
decline words. The latter is done by smart paradigms: functions that analyse
some given forms of a word to find out how the inflection table should look.
For example, the declination of many Swedish nouns can be determined by looking
at the singular form only. 
\begin{wrapfigure}{l}{0.4\textwidth}
\begin{tabular}{| l | l |}
\hline
1st declination & 5th declination \\
\hline
flick\textbf{a}    &     hjärt\textbf{a}   \\
flick\textbf{an}    &    hjärt\textbf{at}  \\
flick\textbf{or}    &    hjärt\textbf{an}  \\
flick\textbf{orna}  &    hjärt\textbf{ana} \\
\hline
\end{tabular}
\end{wrapfigure}
This is the case for `flicka', a noun belonging to the first declination. %the correct declination
%can be guessed by using the first declination. % and noting that the word ends with an `a'. 
%Like most Swedish nouns ending with an `a', it follows the first declination.
For others, like "hjärta", also the plural form "hjärtan" is needed.
%, even though it ends with an `a'. 
The worst case for nouns is four needed forms, both singular and
plural in definite and indefinite form.
Section \ref{sec:swegf} will give a more thorough description of the Swedish resource grammar.

@@

\section{Talbanken}
\label{sec:talbanken}
For testing and evaluation of the grammar and lexicon, we needed to be able to
compare them against a reliable source.
Talbanken \cite{talbanken} was perfect for our purpose,
being a freely available, manually annotated, large-scale treebank.
%It was created in the 1970s at Lund University.
It is analyzed with the MAMBA annotation scheme (Teleman, 1974) and 
consists of four
parts. Two of them are transcriptions of spoken language, one a collection of
text written by high school students, and one, section P,
consists of professionally written Swedish gathered from newspapers, brochures and textbooks.
\\
Talbanken was also used to train the Swedish version of the Malt parser \cite{malt}
and was then redistributed in an updated version,
Talbanken05 \cite{talbanken05}.
%This work was done by Nivre, Hall and Nilsson in 2005.
It is released in Malt\footnote{http://w3.msi.vxu.se/~nivre/research/MaltXML.html} 
and Tiger\footnote{http://www.ims.uni-stuttgart.de/projekte/TIGER/TIGERSearch/doc/html/TigerXML.html}
XML-formats
where the trees have been made deeper and more detailed while still containing
the lexical MAMBA layer. \\
The Malt parser was trained on section P of Talbanken, and these
more than 6000 sentences have been used our project. 
The treebank has served as an inspiration and an evaluation source throughout the
project. An automatic mapping between its trees and the abstract trees from GF has been
done, which will be explained in section \ref{sec:Mapping}.
%Differences in analyse, do we want a similar? 
%Uses of Talbanken in project, mapping to evaluate and test.  
%Look at output from Maltparser, could also comparing our parse trees to this.

\section{Saldo}
\label{sec:saldo}
A good parser needs a good lexicon. We have used Saldo \cite{saldo}, a
large electronic lexicon developed and maintained at Gothenburg University. It is
built on Svenskt Associationslexikon and contains information about more than 
120 000 modern Swedish words.
For each word there is semantic, syntactical and a morphological
information. The user can find examples of usage in corpora,
graphs of semantically connected words and some suggestions for how to analyse compounds. \\
\begin{figure}[h!]
\centering
\subfloat[Morphological info for `katt']{\label{pic:saldoTab}\includegraphics[width=40mm]{saldotab.png}}
\hspace{20mm}
\subfloat[Graph showing the hyponyms of `katt']{\label{pic:saldoWheel}\includegraphics[width=70mm]{saldograph.png}}
\caption{Information in Saldo}
\label{fig:saldo}
\end{figure}
